% intro.tex
\cleardoublepage
\section*{Foreword}
{\large
{
\setlength{\parindent}{2em}
I was reading recently in Romans and noticed a footnote in my Bible that said, ``Romans 3:14 Psalm 10:7 (see Septuagint)''. It got me wondering, just how different \textit{is} the Septuagint? I've done a bit of work with the Greek translation of the Hebrew scriptures, but never really done a side by side comparison. Having recently done some other work that involved putting copies of a couple of individual books side by side, with a focus on where there are known significant differences, I decided to extend that and produce a parallel version of the whole of the Hebrew scriptures in English. This is the result.

The Greek version of the Hebrew scriptures tends to be overlooked in the Western Protestant tradition, but it is worth remembering that although the original text was undoubtedly written in the Hebrew language, the Septuagint was translated from these Hebrew scriptures hundreds of years before the MT was fully settled. Even though it is a translation it provides key insight into the history of the transmission of the text. Where there are differences between the MT and the LXX we should consider that the LXX may actually represent an earlier textual tradition. Indeed, the discovery of the Dead Sea Scrolls supports this with a number of documents in Hebrew that are closer to the LXX than they are to the modern MT.

\subsection*{Sources}
My source text for a standard English Bible, translated almost exclusively from the Hebrew Masoretic Text (MT) that has been the standard source ever since Jerome translated the Latin Vulgate in the early 5th century, I have selected the World English Bible (WEB). This was selected because it has been placed in the public domain, meaning not only am I free to use it, but I am also free to distribute any derived work, like this one. The WEB comes in two flavours, with US or British English. This affects mainly spelling but there are a few other differences. It was simple enough for me to script using either version and as a result this document is also available in two versions. This is the
\ifbritish
British English
\else
US
\fi\ version. The WEB text has not been edited apart from some re-ordering necessary to present parallel passages together. Footnotes and poetry structure reproduce the notes and structure in the  WEB source text. A moderate number of cross references have been added using data from Glenton Jelbert, \url{https://glentonjelbert.com/introducing-the-cross-referenced-bible/}. There are much more comprehensive cross reference databases available but extensive cross referencing was not the purpose of this project.

\ifbritish
The WEB British English edition has frequent repeats of this footnote:\\
\textbf{When rendered in ALL CAPITAL LETTERS, “LORD” or “GOD” is the translation of God’s Proper Name (Hebrew “\texthebrew{ יהוה}”, usually pronounced Yahweh).}; or\\
\textbf{LORD or GOD in all caps is from the Hebrew\texthebrew{ יהוה} Yahweh except when otherwise noted as being from the short form\texthebrew{ יה} Yah.}.\\
The sometimes frequent repetition of the identical footnote has been judged as unnecessary and these have been removed from this edition.
\fi

For the Septuagint I have used Brenton's English translation of the Septuagint. This is a 19th century translation and showing its age, but its age also makes it in the public domain and so free to make derivative works and distribute. (For a more contemporary English translation of the Septuagint, check out the NETS project.) In order to make the text more accessible to a modern reader, I have scripted changes to replace the most archaic language with contemporary equivalents, and manually edited the result with reference to the original Greek. My reference for the original Greek is the 1935 Rahlfs LXX that is mostly based on Codex Vaticanus (B). Brenton's 1851 translation is based on the Sixtine edition (1587) and mostly matches the Rahlfs text but may deviate from it where there are different readings in different Greek manuscripts. This is a work in progress and I will doubtless make further refinements to this as I encounter issues. Words in italics indicating translator's addition of words absent in the original but necessary for the English to make sense are almost all as in Brenton although I have made a small number of additions, deletions and changes to these that are consistent with Brenton's approach. Most footnotes are similarly Brenton's. Brenton's translation does contain errors; I have corrected these where I have spotted them, but my work is mostly to make the translation understandable to a modern reader and not a full revision of Brenton's work so it is likely that some of his errors remain, not to mention any errors that I may have introduced. Text set in small caps and descriptive headings are also as specified by Brenton.

With the exception of Jeremiah (see below) chapter divisions are as they appear in the WEB.

\subsection*{Comments on specific books}
Some books need special comment.

Chapters 36, 37, 38 and 39 of \textbf{Exodus} are quite different in the MT and LXX. I have reordered the LXX a little in an attempt to put parallel verses together. You can reconstruct the original order using the chapter and verse references attached to each verse.

The MT of the books of \textbf{1 and 2 Samuel} is particularly unreliable, and the Old Greek of the LXX has its own issues. I have set a third column in parallel with them that has my translation of the Palestinian text of 1 and 2 Samuel, based on an updated American Standard Version with adjustments made where the Palestinian text deviates from the MT. This column is labelled (P). I am indebted to E.H. Blackmore for the work he did on tabulating the differences between the MT and P texts.

The Greek of the book of \textbf{Esther} is substantially longer than the MT, with some extended additional sections. These additions are generally considered deuterocanonical and are usually separated out in a way that breaks the flow of the book. Esther is presented here with the additions in the order they appear in the Greek text. Additions are identified by a letter in place of a chapter number.

The LXX version of \textbf{Jeremiah} is structured differently to the MT, with the oracles against the nations included in the middle where the MT has them at the end. This leaves us with the problem of what order to present the material here. Because most readers will have access to the order of the MT in their usual English Bible, the decision was made that in this edition Jeremiah would be presented in the order it appears in the LXX. Chapter and verse numbering follows
the Brenton LXX (including a/b suffixes where present). Verse matching between versions was prepared based on the information provided at \url{https://www.ccel.org/bible/brenton/Jeremiah/appendix.html}. 

The LXX version of \textbf{Daniel} is substantially longer than the MT.  The additions have been included in they order they are in the Greek. Unlike most of the Old Testament, where one Greek translation (the Septuagint) dominates, Daniel exists in two different Greek forms. The Old Greek version of Daniel, prepared 1st or 2nd century BC, roughly the same time as the rest of the LXX, was replaced with Theodotian's version around the second century AD. This is a revision of the Greek text to bring it more into conformity with the Hebrew and it does include the additional material that is absent from the Hebrew (and appears to have been originally composed in Greek). This is the version that Brenton translates, and hence is the version presented here as LXX. I present an AI-translated English version of the Old Greek as a third column, labelled (OG), with the translation based on a source obtained from \url{http://ccat.sas.upenn.edu/gopher/text/religion/biblical/lxxmorph/61.DanielOG.mlxx}. In addition to the Hebrew text, the Greek versions include three additions. Two can be considered separate works and are generally considered to be part of the Apocrypha: \textit{Susanna}\/ is sometimes included as Daniel 13, but sometimes comes first in the Greek text since it appears to be set earlier. \textit{Bel and the Dragon}\/ appears at the end of the book, often as Daniel 14. These both are really separate works. The third addition is a major insertion between 3:23 and 3:24, including the \textit{Prayer of Azariah} and the \textit{Song of the Three Holy Children.}\/ This was probably composed separately and later than the rest of Daniel but there is no evidence of it having been circulated as a separate work or works. Since my focus here is not to present the whole of the Apocrypha but to enable comparison between textual traditions of texts that appear in both, I have made the somewhat arbitrary decision not to include \textit{Susanna}\/ or \textit{Bel and the Dragon,}\/ but to leave the chapter 3 insertions in place.

\subsection*{Brenton footnote abbreviations}
As mentioned above, the footnotes to the LXX column are mostly unchanged from Brenton's translations, and any changes have followed his style. He uses the following abbreviations and symbols:\par
\begin{tabular}{ll}
  \textit{Heb.}&for Hebrew.\\
  \textit{Gr.}&for Greek. \\
  \textit{Lit.}&for Literally. \\
  \textit{q. d.}&for \textit{quasi dicat} (as if he were saying).\\
  \textit{Comp.}&for Compare. \\
  \textit{A. V.}&for Authorised Version. \\
  \textit{Alex.}&for Alexandrine Text. \\
  \textit{Ald.}&for Aldine Text.\\
  \textit{App.}&for Appendix. \\
  +&for Sign of addition. \\
  —&for Sign of ommission. \\
  \textit{sc.}&\textit{scilicet}, that is to say.
\end{tabular}\par
}
\bigskip
\noindent Glen Prideaux\\
March 2026

\bigskip
\noindent\textbf{Left:} MT English (WEB)\\
\textbf{Right:} LXX English (Brenton, updated)\\
Where a third column exists it is as noted above and at the beginning of the book.

\cleardoublepage
\chapter*{Contents}
\begin{multicols}{2}
\TOCwithouttitle
\end{multicols}
}